---
title: "STA 561 HW3 (Convexity, Logistic, Boosting)"
author: "Daniel Truver"
date: "2/7/2018"
header-includes:
  - \usepackage{amsmath}
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### (1) Linear Separability Implies Disjoint Convex Hulls

We proceed with proof by contradiction. 

For $\{x_n\}$ and $\{x_m'\}$, let their convex hulls be given by
$$
\begin{aligned}
& X = \{x: x = \sum_n\alpha_nx_x; ~ \alpha_n \geq0; ~\sum_n\alpha_n = 1 \} \\
& X' = \{x:x = \sum_m\alpha_m'x_m';~\alpha_m'\geq0;~\sum_m\alpha_m' = 1 \}
\end{aligned}
$$

(RAA) Suppose $\{x_n\}$ and $\{x_m'\}$ are linearly separable such that for a vector $w$ and scalar $w_0$ $w^Tx_n + w_0 >0,~ w^Tx_m' + w_0 < 0 ~\forall x_n, x_m'$ and $X\cap X' \neq \emptyset$.

Then $$\exists~ x \in X\cap X'$$
such that 
\begin{align}
x &= \sum_n\alpha_nx_n = \sum_m\alpha_m'x_m'\\
w^Tx &= w^T\sum_n\alpha_nx_n = w^T\sum_m\alpha_m'x_m'\\
w^Tx &= \sum_n\alpha_n(w^Tx_n) = \sum_m\alpha_m'(w^Tx_m') 
\end{align}

But, from linear separability, we have $w^Tx_n > -w_0$ and $w^Tx_m' < -w_0$. If we use these relations in the summations above, we obtain
$$
\begin{aligned}
& \sum_n\alpha_n(w^Tx_n) > \sum_n\alpha_n(-w_0), \quad  \sum_m\alpha_m'(w^Tx_m') < \sum_m\alpha_m'(-w_0) \\
& \implies \sum_n\alpha_n(w^Tx_n) > -w_0\sum_n\alpha_n, \quad \sum_m\alpha_m'(w^Tx_m') < -w_0\sum_m\alpha_m' \\
& \implies \sum_n\alpha_n(w^Tx_n) > -w_0, \quad \sum_m\alpha_m'(w^Tx_m') < -w_0 \\
& \implies w^Tx > -w_0 \quad\text{and}\quad w^Tx < -w_0 \quad\text{by (3)} \\
& ~\Rightarrow\!\Leftarrow
\end{aligned}
$$

Therefore, $X$ and $X'$ are disjoint; the convex hulls do not intersect. 

******

#### (2) Logistic Regression and Gradient Descent

##### (a) Logistic Sigmoid Function 

$$
\begin{aligned}
\sigma'(a) 
&= \frac{d}{da}(1+e^{-a})^{-1} \\
&= e^{-a}(1+e^{-a})^{-2} \\
&= (1+e^{-a})^{-1}\left(\frac{e^{-a}}{1+e^{-a}}  \right) \\
&= \sigma(a)\left( \frac{1+e^{-a}-1}{1+e^{-a}} \right) \\
& = \sigma(a)\left( 1- \frac{1}{1+e^{-a}} \right) \\
&= \sigma(a)(1-\sigma(a))
\end{aligned}
$$  

##### (b) Derivative of Cross Entropy 

$$
\begin{aligned}
\frac{\partial}{\partial w_j}L_w
&= \sum_{i = 1}^n \frac{\partial}{\partial w_j}\left(-y^{(i)}\log\sigma(w^Tx^{(i)})\right) - \frac{\partial}{\partial w_j}\left( (1- y^{(i)})\log(1-\sigma(w^Tx^{(i)})) \right) \\
&= \sum_{i = 1}^n -y^{(i)}\frac{1}{\sigma(w^Tx^{(i)})}\sigma(w^Tx^{(i)})(1-\sigma(w^Tx^{(i)}))(-x_j^{(i)})\\
&\quad \quad \quad - (1-y^{(i)})\frac{1}{1-\sigma(w^Tx^{(i)})}\left(-\sigma(w^Tx^{(i)})\right)\left(1-\sigma(w^T)\right)(-x_j^{(i)}) \\
&= \sum_{i=1}^n y^{(i)}x_j^{(i)}\left( 1-\sigma(w^Tx^{(i)}) \right) - (1-y^{(i)})x_j^{(i)}\sigma(w^Tx^{(i)}) \\
&= \sum_{i = 1}^n x_j^{(i)}\left( y^{(i)} \big( 1-\sigma(w^Tx^{(i)}) \big) - (1-y^{(i)})\sigma(w^Tx^{(i)}) \right)
\end{aligned}
$$  

##### (c) Convexity of the Cross Entropy Loss 

We want to show that, for $x, z \in \mathbb{R}^p$ and $\theta \in [0,1]$, 
$$
L_w(\theta x + (1-\theta)z) \leq \theta L(x) + (1-\theta)L(z).
$$

It suffices to show that $\forall i$, 
$$
g(x^{(i)}) = -y^{(i)}\log\sigma(w^Tx^{(i)}) -(1- y^{(i)})\log(1-\sigma(w^Tx^{(i)}))
$$
is convex.

We consider 2 cases, $y^{(i)}=1$ and $y^{(i)} =0$. 
Case 1 ($y^{(i)} =1$):
$$
g(x) = -\log\sigma(w^Tx) = \log(1+e^{-w^Tx})
$$

If we take the second derivative of this function in any direction $x_j$, we get
$$
\begin{aligned}
\frac{\partial^2}{\partial^2x_j}g(x) 
&= \frac{\partial}{\partial x_j}\left( -w_j e^{-w^Tx}(1+e^{-w^Tx})^{-1} \right) \\
&= w_j^2 e^{-w^Tx}(1+e^{-w^Tx})^{-1} - w_j^2(e^{-w^Tx})^2(1+e^{-w^Tx})^{-2} \\
&= w_j^2 e^{-w^Tx}(1+e^{-w^Tx})^{-1} (1- e^{-w^Tx}(1+e^{-w^Tx})^{-1}) \\
&\geq 0
\end{aligned}
$$

Therefore, this function is convex. 

Case 2 ($y^{(i)} = 0$):
$$
f(x) = -\log(1-\sigma(w^Tx))
$$
Taking the second derivative w.r.t. $x_j$, we get 
$$
\begin{aligned}
\frac{\partial^2}{\partial^2x_j}f(x) 
&= \frac{\partial^2}{\partial^2x_j} \left( -\log(e^{-w^Tx}) -\log(1+e^{-w^Tx})^{-1} \right)  \\
&= \frac{\partial}{\partial x_j}\left( w_j -w_j e^{-w^Tx}(1+e^{-w^Tx})^{-1} \right) \\
&\geq 0
\end{aligned}
$$
since this derivative will be the same as Case 1. 

Therefore, we have shown the summand of the cross entropy loss function to be complex $\forall i$. The positive sum of convex functions is convex, so the cross entropy loss is convex. 

##### (d) See Code

******

#### (3) Boosting

##### (a) Perfect Training Classification 

Assume the the weak learning assumption (WLA). So, we get 
$$
\epsilon_t = \text{error}_{d_t}(h_{(t)}) =: \frac{1}{2} - \gamma_t, \quad \gamma_t > \gamma_{WLA}
$$

Now sit and stare at the problem for a couple hours. Get frustrated. Eat a sandwich. Come back to the problem. Write down thought process.

```{r allworkandnoplay, echo=FALSE}
tootrue = rep("\nAll work and no play makes Jack a dull boy.",5)
cat(tootrue)
```  

Another sandwich. Scroll up in the lecture notes. Epiphany.

$$
\begin{aligned}
\text{classification error}
&\leq \frac{1}{n}\sum_{i=1}^n e^{-y_if(x_i)}\\
&= \frac{1}{n} \sum_{i=1}^n \exp\left( -y_i\sum_{t=1}^T\alpha_th_{(t)}(x_i) \right) \\
&= \frac{1}{n} \sum_{i=1}^n\prod_{t=1}^T \exp\left( -y_i\alpha_th_{(t)}(x_i) \right) \\
&= \sum_{i=1}^n \frac{1}{n}\exp\left( -y_i\alpha_1h_{(1)}(x_i) \right)\cdots\exp\left( -y_i\alpha_Th_{(T)}(x_i) \right) \\
&= \sum_{i=1}^n d_{1,i}\exp\left( -y_i\alpha_1h_{(1)}(x_i) \right)\cdots\exp\left( -y_i\alpha_Th_{(T)}(x_i) \right) \\
&= \sum_{i = 1}^n d_{2,i}Z_1\exp\left( -y_i\alpha_2h_{(2)}(x_i) \right)\cdots\exp\left( -y_i\alpha_Th_{(T)}(x_i) \right) \\
&= \sum_{i=1}^n d_{T+1,i}(Z_1\cdots Z_T) \\
&= \prod_{t=1}^T Z_t \quad \quad (d_{T+1} \text{ is a distribution}) 
\end{aligned}
$$

We now have the error bounded by the product of the normalization constants which appear to be functions of $t$. We now want to find a bound of each indivual $Z_t$, hopefully still a function of $t$, and show that everything together approaches 0 as $t$ approaches $\infty$. 

$$
\begin{aligned}
Z_t 
&= \sum_{i=1}^n d_{t,i} e^{-y_i\alpha_th_{(t)}(x_i)} \\
&= \sum_{i:h_t(x_i) = y_i} d_{t,i}e^{-\alpha_t} + \sum_{i:h_t(x_i) \neq y_i} d_{t,i}e^{\alpha_t}\\
&= (\gamma_t+1/2)e^{-\alpha_t} + (1/2-\gamma_t)e^{\alpha_t} \quad \quad (\text{by WLA}) \\
&= (\gamma_t+1/2)\left(\frac{1/2-\gamma_t}{\gamma_t+1/2}\right)^\frac{1}{2} + (1/2-\gamma_t)\left(\frac{\gamma_t+1/2}{1/2-\gamma_t}\right)^\frac{1}{2}\quad (\text{by defn. of } \alpha_t) \\
&= 2\sqrt{(\gamma_t+1/2)(1/2-\gamma_t)} \\
&= \sqrt{1 - 4\gamma_t^2} \\
&< \sqrt{1-4\gamma_{WLA}^2} \quad\quad (\gamma_t > \gamma_{WLA}) \\
&= r < 1 \quad \quad \left(\gamma_{WLA}\in(0,\frac{1}{2})\right)
\end{aligned}
$$

This gives us 
$$
\text{training error} < \prod_{t=1}^T Z_t < \prod_{t=1}^T r = r^T
$$

Since $r<1$, we can view $r^T$ as a monotonically decreasing sequence in $T$. Given $\epsilon > 0$, take
$$T > \frac{\log{\epsilon}}{\log{r}}$$
Then we have 
$$
T\log r < \log\epsilon \implies r^T<\epsilon
$$

Bringing it all together, we get $\forall \epsilon > 0, \exists ~T$ such that.
$$
\text{training error} < \epsilon 
$$

##### (b) Weighted Points in Training Set

Suppose that our objective is
$$
R^{train}(\lambda) = \sum_{i=1}^n w_ie^{-(M\lambda)_i}
$$

It seems that the standard AdaBoost we already derived is just a special case of this where all weights are $w_i = 1/n$. For this reason, the derivation and results should look similar. We are still going to look for an optimal direction and step size in that direction to iteratively reduce the objective above. That is, we are still doing coordinate descent, just with more subscripts to track.

$$
\begin{aligned}
j_t
&\in \text{argmax}_j\left[ -\frac{\partial R^{train}(\lambda_t + \alpha \mathbf{e}_j)}{\partial\alpha}\Biggr|_{\alpha = 0} \right] \\
&= \text{argmax}_j\left[ -\frac{\partial}{\partial\alpha} \left[ \sum_{i=1}^n w_ie^{-(M(\lambda_t + \alpha e_j))_i} \right] \Biggr|_{\alpha = 0} \right] \\
&= \text{argmax}_j\left[ \sum_{i=1}^n w_iM_{ij}e^{-(M\lambda_t)_i} \right] \quad \text{(some algebra later)}
\end{aligned}
$$

This time, our discrete probability distribution will be 
$$
d_{t,i} = w_ie^{-(M\lambda_t)_i}/Z_t \text{ where } Z_t = \sum_{i=1}^n w_ie^{-(M\lambda_t)_i} 
$$

Same as before, the argmax does not change when we multiply by $Z_t$ and we have absorbed the $w_i$ into our distribution, so we are left with 
$$
j_t \in \text{argmax}_j\sum_{i=1}^n M_{ij}d_{t,i} = \text{argmax}(d_t^TM)_j.
$$  

Our caculation of the step size is also fairly similar. We search along the direction $j_t$ by setting the derivative equal to 0. 
$$
\begin{aligned}
0 
&= \frac{\partial R^{train}(\lambda_t+\alpha e_{j_t})}{\partial\alpha}\Biggr|_{\alpha_t}\\
&= -\sum_{i=1}^n w_i M_{ij_t} e^{-(M\lambda_t)_i-\alpha_tM_{ij_t}} \\
&= -\sum_{i:M_{ij_t = 1}}w_i e^{-(M\lambda_t)_i}e^{-\alpha_t} - \sum_{i:M_{ij_t = -1}w_i - e^{(M\lambda_t)_i}e^{\alpha_t}
\end{aligned}
$$
 
Now we multiply by $-1/Z_t$ to get us one step closer to the end of our suffering:

$$
\begin{aligned}
0 &= \sum_{i:M_{ij_t = 1}} d_{t,i}e^{-\alpha_t} - \sum_{i:M_{ij_t = -1}}d_{t,i}e^{\alpha_t}\\
&=: d_+e^{-\alpha_t} - d_-e^{\alpha_t} \\
d_-e^{\alpha_t} &= d_+e^{-\alpha_t} \\
\alpha_t &= \frac{1}{2}\ln\frac{1-d_-}{d_-}
\end{aligned}
$$

So, in the end, the coordinate descent algorithm is...exactly the same. The only difference is that 
$$
d_{1,i} = w_i \text{ for } i = 1,\dots, n.
$$